---
title: "basic descriptions"
author: "YX"
date: "2026-01-31"
output: html_document
---

Next step: add loops for each year 2021-2025, need to figure out how the years differ 
Next step: make datasets unique to day, month, year
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

0. Load packages 
```{r}
library(tidyverse)
library(timeDate) # for US federal holidays
```

1. Load data 
```{r, warning = F}
# Annual data
# getwd()
setwd("/Users/yungexiao/Desktop/bluebike analysis")
# rides2021 <- read.csv("rides_2021_adj.csv") # no rideable type
rides2022 <- read.csv("processed/rides_2022_adj.csv") # no rideable type 
# rides2023 <- read.csv("rides_2023.csv")
# rides2024 <- read.csv("rides_2024.csv")
# rides2025 <- read.csv("rides_2025.csv")
stations <- read.csv("data/stations.csv")
```

2. First brief looks
```{r}
# glimpse(rides2022)
# glimpse(stations)

colnames(rides2022) # check variable names 
```

2.1 Check for duplicate stations before merging
```{r}
length(unique(c(rides2022$start_station_name, rides2022$end_station_name))) # unique stations in ride data
length(stations$station_name) # 597 unique stations in stations data

# stations %>% 
  # group_by(station_id) %>% 
  # summarise(count = n()) %>% 
  # filter(count > 1) # id 214 is duplicated

station_dup <- stations %>% 
  filter(station_id == "214") %>% # 2 stations with same ID (214) but different station number
  select(station_id) %>% 
  distinct() 

dup_id = station_dup$station_id
```

```{r}
# check which version of the duplicate is represented in the data 
station_tokeep <- rides2022 %>% 
  filter(start_station_id == 214 | start_station_id == "Bremen St at Marion St" | start_station_id == "Airport T Stop - Bremen St at Brooks St") %>% 
  select(start_station_name, start_station_id) %>% 
  distinct() 

keep = station_tokeep$start_station_name # store station to keep 

# de-duplicate stations correctly 
stations_tomerge <- stations %>% 
  filter(!(!(station_name %in% keep) & station_id %in% dup_id)) %>% # remove correct duplicate
  select(-X, -station_number, -station_lat, -station_long, -station_number) # select necesary variables for merge
```

```{r}
rides2022 <- rides2022 %>% 
  mutate(station_id = as.character(start_station_id),
         end_station_id = as.character(end_station_id), # convert to character for merge
         started_at = ymd_hms(started_at),
         ended_at = ymd_hms(ended_at)) %>% # convert to datetime
  left_join(stations_tomerge, by = "station_id") %>% 
  select(-station_id, -start_station_id) %>% 
  rename(start_municipality = station_municipality,
         station_id = end_station_id) %>% 
  left_join(stations_tomerge, by = "station_id") %>% 
  rename(end_municipality = station_municipality) %>% 
  select(ride_id, rideable_type, started_at, ended_at, start_station_name, end_station_name, start_municipality, end_municipality, member_casual) 

colnames(rides2022)
```

2.2 Add key date/time variables and flags 
```{r}
rides2022 <- rides2022 %>% 
  mutate(start_date = date(started_at),
         # end_date = date(ended_at),
         start_month = month(started_at),
         # end_month = month(ended_at),
         start_day = day(started_at),
         # end_day = day(ended_at),
         start_week = week(started_at),
         # end_week = week(ended_at),
         start_year = year(started_at),
         # end_year = year(ended_at),
         start_hour = hour(started_at),
         # end_hour = hour(ended_at),
         start_dow = weekdays(started_at),
         # end_dow = weekdays(ended_at)
         ) %>% 
  mutate(wkday_flag = ifelse(start_dow %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"), 1, 0),
         season = case_when(
                    start_month == 12 & start_day >= 21 | start_month %in% c(1, 2) |
                      (start_month == 3 & start_day < 20) ~ "winter",
                    start_month == 3 & start_day >= 20 | start_month %in% c(4, 5) |
                      (start_month == 6 & start_day < 21) ~ "spring",
                    start_month == 6 & start_day >= 21 | start_month %in% c(7, 8) |
                      (start_month == 9 & start_day < 22) ~ "summer",
                    start_month == 9 & start_day >= 22 | start_month %in% c(10, 11) |
                      (start_month == 12 & start_day < 21) ~ "autumn")
         ) %>% 
  mutate(ride_dur_sec =as.numeric(difftime(ended_at, started_at)), # ride duration -- seconds
         ride_dur_min = as.numeric(difftime(ended_at, started_at, units = "mins"))) # ride duration -- minutes

# check flags 
rides2022 %>% 
  ggplot(aes(x = start_dow, fill = as.factor(wkday_flag))) + geom_bar() # looks right

rides2022 %>% 
  ggplot(aes(x = as.factor(start_month), fill = season)) + geom_bar() # looks right

# check duration 
summary(rides2022$ride_dur_min) # very right skewed
# quantile(rides2022$ride_dur_min, probs = seq(0, 1, 0.05)) # check quantiles 
dur95 = quantile(rides2022$ride_dur_min, probs = 0.95) # store 95th percentile of ride duration 

rides2022 %>% 
  filter(ride_dur_min <= dur95) %>% 
  ggplot(aes(x = ride_dur_min)) + geom_boxplot() + 
  labs(title = "ride duration distribution for durations <= 95th %-tile") # a lot of high outliers 
```
Save data
```{r}
setwd("/Users/yungexiao/Desktop/bluebike analysis")
write.csv(rides2022, "processed/rides_2022_adj_1.csv")
```

2.3 Basic trends over time -- number of rides and ride duration 

2.3.0 Number of rides and median duration per ride
```{r}
# overall 
rides2022 %>% 
  summarise(n_rides = n(),
            med_dur_min = median(ride_dur_min)) 
```

```{r}
# per day 
rides_pday <- rides2022 %>% 
  group_by(start_date) %>% 
  summarise(n_rides = n(),
            med_dur_min = median(ride_dur_min)) 

# per hour of day 
rides_phour <- rides2022 %>% 
  group_by(start_hour) %>% 
  summarise(n_rides = n(),
            med_dur_min = median(ride_dur_min)) 

# per week 
rides_pweek <- rides2022 %>% 
  group_by(start_week) %>% 
  summarise(n_rides = n(),
            med_dur_min = median(ride_dur_min)) 

# per month 
rides_pmonth <- rides2022 %>% 
  group_by(start_month) %>% 
  summarise(n_rides = n(),
            med_dur_min = median(ride_dur_min)) 
```

```{r}
# per day 
rides_pday %>% ggplot(aes(x=start_date, y = n_rides)) + geom_point()
rides_pday %>% ggplot(aes(x=start_date, y = med_dur_min)) + geom_point()

# per hour of day 
rides_phour %>% ggplot(aes(x=start_hour, y = n_rides)) + geom_point()
rides_phour %>% ggplot(aes(x=start_hour, y = med_dur_min)) + geom_point()

# per week 
rides_pweek %>% ggplot(aes(x=start_week, y = n_rides)) + geom_point()
rides_pweek %>% ggplot(aes(x=start_week, y = med_dur_min)) + geom_point()

# per month 
rides_pmonth %>% ggplot(aes(x=start_month, y = n_rides)) + geom_point()
rides_pmonth %>% ggplot(aes(x=start_month, y = med_dur_min)) + geom_point()
```

2.3.1 Number of rides and median duration across groups
```{r}
# per season 
rides_pszn <- rides2022 %>% 
  group_by(season) %>% 
  summarise(n_rides = n(),
            med_dur_min = median(ride_dur_min)) 

# per type of bike -- missing for 2021-2022  
rides_ptype <- rides2022 %>% 
  group_by(rideable_type) %>% 
  summarise(n_rides = n(),
            med_dur_min = median(ride_dur_min)) 

# per member type 
rides_pmemtype <- rides2022 %>% 
  group_by(member_casual) %>% 
  summarise(n_rides = n(),
            med_dur_min = median(ride_dur_min)) 

# per dow 
rides_pdow <- rides2022 %>% 
  group_by(start_dow) %>% 
  summarise(n_rides = n(),
            med_dur_min = median(ride_dur_min)) 

# per weekday/weekend status
rides_pwkday <- rides2022 %>% 
  group_by(wkday_flag) %>% 
  mutate(wkday_flag = as.factor(wkday_flag)) %>% 
  summarise(n_rides = n(),
            med_dur_min = median(ride_dur_min)) # 1 is weekday, 0 is weekend
```

```{r}
# per season 
rides_pszn %>% ggplot(aes(x= season, y = n_rides, fill = season)) + geom_col()
rides_pszn %>%  ggplot(aes(x = season, y = med_dur_min, fill = season)) + geom_col()

# per type of bike -- missing for 2021-2022  
rides_ptype %>% ggplot(aes(x= rideable_type, y = n_rides, fill = rideable_type)) + geom_col()
rides_ptype %>%  ggplot(aes(x = rideable_type, y = med_dur_min, fill = rideable_type)) + geom_col()

# per member type 
rides_pmemtype %>% ggplot(aes(x= member_casual, y = n_rides, fill = member_casual)) + geom_col()
rides_pmemtype %>%  ggplot(aes(x = member_casual, y = med_dur_min, fill = member_casual)) + geom_col()

# per dow 
rides_pdow %>% ggplot(aes(x= start_dow, y = n_rides, fill = start_dow)) + geom_col()
rides_pdow %>%  ggplot(aes(x = start_dow, y = med_dur_min, fill = start_dow)) + geom_col()

# per weekday/weekend status
rides_pwkday %>% ggplot(aes(x= wkday_flag, y = n_rides, fill = wkday_flag)) + geom_col()
rides_pwkday %>%  ggplot(aes(x = wkday_flag, y = med_dur_min, fill = wkday_flag)) + geom_col()
```

```{r}
# per municipality -- maybe make long if we don't care about start/stop difference
rides_pstartmun <- rides2022 %>% 
  group_by(start_municipality) %>% 
  summarise(n_rides = n(),
            med_dur_min = median(ride_dur_min)) 

rides_pendmun <- rides2022 %>% 
  group_by(end_municipality) %>% 
  summarise(n_rides = n(),
            med_dur_min = median(ride_dur_min)) 

# per station -- maybe make long
rides_pstartstation <- rides2022 %>% 
  group_by(start_station_name) %>% 
  summarise(n_rides = n(),
            med_dur_min = median(ride_dur_min)) 

rides_pendstation <- rides2022 %>% 
  group_by(end_station_name) %>% 
  summarise(n_rides = n(),
            med_dur_min = median(ride_dur_min)) 
```

2.3.2 Number of rides and median ride duration by group status -- maybe make this a different rmarkdown file












